Arrows for Parallel Computations

### 2017-09-26 paper/content/parrows.tex
minor update
replaced "Multicore Haskell" with "GpH"

\section{Parallel Arrows}
\label{sec:parallel-arrows}
Arrows are a general interface to computation. Here we introduce special Arrows as general interface to \textit{parallel computations}. First, we present the interface and explain the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we also introduce an additional |conf| parameter to the function. We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
We do not require the |conf| parameter in every implementation. If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
% , as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.

@@

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq))
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.

@@

\subsubsection{Eden}
For both the Multicore Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{code}
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{code}
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
% We don't do it here
However, we avoid doing so for aesthetic resons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
% \begin{figure}[h]
\begin{code}
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{code}
% \caption{ArrowParallel instance for functions in the Eden backend}
% \label{fig:ArrowParallelEdenFns}
% \end{figure}
and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
% \begin{figure}[h]
\begin{code}
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{code}
% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
% \label{fig:ArrowParallelEdenKleisli}
% \end{figure}

>>>>

\section{Parallel Arrows}
\label{sec:parallel-arrows}
Arrows are a general interface to computation. Here we introduce special Arrows as general interface to \textit{parallel computations}. First, we present the interface and explain the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we also introduce an additional |conf| parameter to the function. We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
We do not require the |conf| parameter in every implementation. If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
% , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.

@@

\subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
The GpH implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq))
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the GpH Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the GpH interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|. With this definition in place, we can provide a delegating version of the non-configurable instance. We show this in Fig.~\ref{fig:actualNonConfigMulticore}.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the GpH Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.

@@

\subsubsection{Eden}
For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{code}
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{code}
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
% We don't do it here
However, we avoid doing so for aesthetic resons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
% \begin{figure}[h]
\begin{code}
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{code}
% \caption{ArrowParallel instance for functions in the Eden backend}
% \label{fig:ArrowParallelEdenFns}
% \end{figure}
and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
% \begin{figure}[h]
\begin{code}
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{code}
% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
% \label{fig:ArrowParallelEdenKleisli}
% \end{figure}

### 2017-09-26 paper/content/topologySkeletons.tex
replaced "Multicore Haskell" with "GpH"

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the Multicore and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and Multicore backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}

>>>>

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}
